The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-11 paper/ms.tex
finished description of the simulation

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121~}
\shorttitle{Bullet Groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

We estimate the expected distribution of displacements between the two
dominant density peaks in dark matter halos simulated in a full
cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. The study extends previous theoretical studies by
simulating two samples: groups and clusters. We find that 50\% to 60\%
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$133\pm21$\hkpc observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $80\%$ to $85\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1\%$ to
$1.0\%$ of the low mass systems should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for halos with circular velocities larger than $700\kms$
this fraction is in the range of $4\%$ to $10\%$, consistent with
previous studies of dark matter to baryon separations in clusters. The
extension of the theoretical predictions and observational results
towards low mass Bullet-like configurations, i.e. larger abundance of
systems,  opens up the possibility for a new statistical test of
$\Lambda$CDM.

\end{abstract}

@@

This paper is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results and observational
perspectives to finally conclude in Sections \ref{sec:conclusions}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250 comoving \hMpc on a side. The simulation uses the ART code to
follow the evolution of  a dark matter density field sampled with
$2048^3$ from $z=80$ to $z=0$ [...]. The cosmology used  corresponds
to  the spatially flat concordance model with the following
parameters:  the density parameter for matter (dark matter$+$baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  The number
of particles used for each of the DM component was $2048^3$, resulting
in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$. \citet{2011ApJ...740..102K}.

We use halo catalogs constructed using the BDM algorithm. [...] 
All the raw data used in this Letter is available through the
Multidark database \footnote{\texttt{www.multidark.org}}
\citep{Riebe2013}.

In the snapshots at redshift $z=0.0,0.25, 0.5$ and $1.0$ there are
$10041, 10346, 10554,$ and $10382$ host halos with circular velocities
$V_{\rm c}\geq 300$\kms; and $157853, 177331, 195072, 225188$
sub-halos with circular velocities $V_{\rm c}\geq 75$\kms,
respectively. These two sets of halos constitute the basis for our
analysis.

For each host halo we find its most massive sub-halo. Each pair
host/sub-halo is considered as a potential Bullet system and is kept
got the analysis described in the next section.

@@

\subsection{Reproducibility of our results}
In order to facilitate the reproducibility and reuse of our results we
have made available all the data and the source code available in a public
repository\footnote{\texttt{https://github.com/Fernandez-Trincado/Bullet\_Groups-2014}.}

>>>>

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121}
\shorttitle{Bullet Groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. We extend previous results by
studying two halo samples: groups and clusters. We find that 50\% to 60\%
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$133\pm21$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $80\%$ to $85\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1\%$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction is in the range of $4\%$ to
$10\%$, consistent with previous studies of dark matter to baryon
separations. The extension of the theoretical predictions and
observational results towards low mass Bullet-like configurations,
i.e. larger abundance of systems,  opens up the possibility for a new
statistical test of  $\Lambda$CDM.  
\end{abstract}

@@

This Letter is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results and observational
perspectives to finally conclude in Sections \ref{sec:conclusions}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side. The simulation uses the ART code to
follow the evolution of a dark matter density field sampled with
$2048^3$ from $z=80$ to $z=0$. The cosmology used  corresponds
to  the spatially flat concordance model with the following
parameters:  the density parameter for matter (dark matter and baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

The number of particles used for each of the DM component was
$2048^3$, resulting in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$. The completeness limit in this simulation is set
for halos with $100$ particles corresponding to a mass of
$1.35\times10^{10}$\hMsun or a maximum circular velocity $V_{c}$ of
$50$\kms.

We use halo catalogs constructed using the Bound Density Maxima (BDM)
algorithm \citep{BDM,BDMb}. To define the radius of a halo we use a
density threshold of 360 times the mean density of the Universe, for 
different redshifts we use the overdensity criterion by
\cite{Bryan1998}. An important feature of BDM is that it allows us to
detect sub-halos inside larger virialized structures.

All the raw data used in this Letter is
available through the Multidark database \footnote{\texttt{www.multidark.org}}
\citep{Riebe2013}.  Furthermore, in order to facilitate the
reproducibility and reuse of our results we have made available all
the data and the source code available in a public
repository \footnote{\texttt{https://github.com/Fernandez-Trincado/Bullet\_Groups-2014}}.

In this Letter we make a study at four different redhifts $z=0.0,0.25,
0.5$ and $1.0$. First, we select all the host halos (i.e. halos that
are not inside a larger halo) with circular velocities $V_{\rm c}\geq
300$\kms. Then, we select the sub-halos with circular velocities
$V_{\rm c}\geq 75$\kms. The objective is to use this sub-halos as the
tracer of the sub-dominant dark matter clump in the merging cluster,
i.e. the bullets.

For the redshifts $z=0.0,0.25,
0.5$ and $1.0$ we find $10041, 10346, 10554$ and $10382$ host halos
and $157853, 177331, 195072, 225188$ sub-halos, respectively. These
two sets (host halos and sub-halos) constitute the basis for our
analysis. We find now for each host halo we its most massive sub-halo. Each pair
host halo/sub-halo is considered as a potential Bullet system and is
kept for the analysis described in the next Section.

@@
